Dependent Object Types

### 2012-08-31 doc/candidacy/report.tex
first draft of my work sec

% TODO
%Describe your own work and include a  summary and references. Write how you
%propose to advance the state of the art given the background. What
%is new technically? How does it improve over prior work?
%Summarize, suggest an approximate timeline, and list references.

>>>>

In our own research, we have been exploring a new core calculus for
path-dependent types, the Dependent Object Types (DOT) calculus, which
could serve as a new type-theoretic foundation for Scala and languages
like it. The three calculi presented demonstrate that path-dependent
types arise in diverse settings, and thus, it is valuable to identify
a core calculus which captures their essence. In DOT, we are not
concerned with modeling orthogonal concepts such as inheritance and
mixin composition. In contrast to the three previous calculi,
type-checking in DOT is decidable, since all the judgments are
syntax-directed (in the current design).

DOT aims to bring more uniformity and simplicity to Scala. Scala, and
${\nu}Obj$ which models it, conflate the concepts of compound types
(which inherit the members of several parent types) and mixin
composition (which build classes from other classes and traits). At
first glance, this offers an economy of concepts. However, it is
problematic because mixin composition and intersection types have
quite different properties. In particular, we have seen in
section~\ref{sec:nuObj} that mixin composition is not commutative,
unlike classical intersection types.

In DOT, we replace Scala's compound types by classical intersection
types, which are commutative. We also complement this by classical
union types. Intersections and unions form a lattice wrt
subtyping. This addresses another problematic feature of Scala: In
Scala's current type system, least upper bounds and greatest lower
bounds do not always exist. Here is an example: Given two traits
\begin{lstlisting}
  trait A { type T <: A }
  trait B { type T <: B }
\end{lstlisting}
The greatest lower bound of \code@A@ and \code@B@ is approximated by the
infinite sequence
\begin{lstlisting}
  A with B { type T <: A with B { type T <: A with B {
    type T < ...
  }}}
\end{lstlisting}
The limit of this sequence does not exist as a type in Scala. This is
problematic because glbs and lubs play a central role in Scala's type
inference. The absence of universal glbs and lubs makes type inference
more brittle and more unpredictable.

Our immediate goal is to prove type soundness of the DOT calculus. We
first tried to prove the calculus type-safe using the standard
theorems of preservation and progress~\cite{soundness,
  tapl}. Unfortunately, for the DOT calculus and any variants that we
devise, preservation doesn't hold. Indeed, we found many
counterexamples to preservation due to narrowing, i.e.\ a type becoming more
precise after substitution. However, the standard theorems of
preservation and progress are just one way to prove type soundness. We
are exploring a promising avenue to prove type safey using the
powerful method of step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In the process of
proving the calculus sound, we are also refining it.

In future work, we plan to investigate translations of Scala features
into the core DOT calculus, and implement an experimental compiler
front-end which uses the DOT calculus for type-checking.

%\clearpage

%Describe your own work and include a  summary and references. Write how you
%propose to advance the state of the art given the background. What
%is new technically? How does it improve over prior work?
%Summarize, suggest an approximate timeline, and list references.
